---
title: "disease_overview"
author: "Stefanie Mueller"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
    lightbox: true
    gallery: true
    toc_depth: 3
    use_bookdown: true
    fig_caption: true
    code_folding: hide
editor_options: 
  chunk_output_type: console
---


```{r knitr_init, echo=FALSE, cache=FALSE}
# load libraries
library(knitr)
library(rmdformats)

# define functions
myDT = function(data){
  data %>% DT::datatable(rownames = FALSE,
    filter = 'bottom',
    extensions = 'Buttons',
      options = list(
        searching = TRUE,
        fixedColumns = TRUE,
        autoWidth = TRUE,
        dom = 'Bfrtip',
        buttons = c('copy', 'csv', 'excel'))
    )
}

## Global options
#options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=TRUE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE,
               fig.width = 10)
opts_knit$set(width=75)
```

# PREFACE
**PROJECT**:  Phenotype Risk Scores  
**OBJECTIVE**:  Summary stats over analysed diseases   
**DEPENDENCE**:  phers.R    
**REQUIREMENT**:  R 3.5.0   
<br>

## Load packages

```{r}

```

```{r}
library(tidyverse)
library(purrr)
library(rstatix)
```

## Script Variables
```{r}
OUTPATH = "~/health_records/smueller/phers/tables/"
SCRIPT = "DisOver"
DATE = Sys.Date()
```

# Load sample data per disease
```{r, eval=T}
#Phenotype Risk Scores per disease
file_list = list.files("/home/smueller/health_records/smueller/phers/samples",
                        pattern="samples_ccratio_4.csv$")

sample_info_LIST = list()

#loop over diseases
for (i in 1:length(file_list)){
  
  disease = str_replace(file_list[i],pattern="_samples_ccratio_4.csv", replacement = "" ) 
  
  # read sample info
  sample_info_LIST[[i]] = read.csv(paste0("/home/smueller/health_records/smueller/phers/samples/",(file_list[i]))) %>% 
    mutate(disease = disease) 
}
  
sample_info = map_dfr(sample_info_LIST, rbind)
```

# Look at Age of Onset

```{r}
#summarise info per disease and particiapnt type (ie. control, case with PRS "0" or case)
sum.dis = sample_info %>% 
  filter(Pheno!="Control") %>% 
  group_by(disease) %>% 
  summarise(Number =n(),
    Median_age = median(age),
    IQR_age = quantile(age,probs = 0.75)-quantile(age,probs = 0.25),
    Median_age_of_onset = median(age_of_onset, na.rm = T),
    IQR_age_of_onset = quantile(age_of_onset, probs=0.75,na.rm = T) - 
      quantile(age_of_onset, probs=0.25,na.rm = T),
    Num_age_of_onset_missing = sum(is.na(age_of_onset)),
    Prop_age_of_onset_missing = Num_age_of_onset_missing/Number,
    Median_encounters = median(Number_uniq_encoutner),
    IQR_encounters = quantile(Number_uniq_encoutner,probs = 0.75)-quantile(Number_uniq_encoutner,probs = 0.25),
    Median_Record_Length = median(Record_Length),
    IQR_Record_Length = quantile(Record_Length,probs = 0.75)-quantile(Record_Length,probs = 0.25),
    Median_Number_uniq_ICD = median(Number_uniq_ICD),
    IQR_Number_uniq_ICD = quantile(Number_uniq_ICD,probs = 0.75)-quantile(Number_uniq_ICD,probs = 0.25)) 


# save
#write_csv(sum.phers_and_pheno, 
#          paste0(OUTPATH,"suppl_table.",SCRIPT,".summary_stat.",DATE,".csv"))
```

## Plot histogram of age of onset


```{r,eval=T}
options(bitmapType="cairo")
X11(type = 'cairo')

sum.dis %>% 
  ggplot(aes(Median_age_of_onset)) + 
  geom_histogram()
```

## Table
```{r}
sum.dis %>% 
  select(disease, Median_age_of_onset, Prop_age_of_onset_missing) %>% 
  mutate(Age_Classifier =ifelse(Median_age_of_onset==0, "At birth",
                         ifelse(Median_age_of_onset<18, "Before 18",
                         ifelse(Median_age_of_onset>60,"Over 60",
                         ifelse(Median_age_of_onset>40,"Over 40", "Between 18 and 40"))))) %>% 
  myDT()
```

```{r}
sum.dis %>% 
  select(disease, Median_age_of_onset, Prop_age_of_onset_missing) %>% 
  mutate(Age_Classifier =ifelse(Median_age_of_onset==0, "At birth",
                         ifelse(Median_age_of_onset<18, "Before 18",
                         ifelse(Median_age_of_onset>60,"Over 60",
                         ifelse(Median_age_of_onset>40,"Over 40", "Between 18 and 40"))))) %>% 
  count(Age_Classifier)
```


# Combine statistical results
```{r}
file_list2 = list.files(path="/home/smueller/health_records/smueller/phers/stat" ,pattern="assoc.stat.csv")

stat_LIST = list()
for(i in 1:length(file_list2)){
  
  stat_LIST[[i]] = read.csv(paste0("/home/smueller/health_records/smueller/phers/stat/",file_list2[i])) %>% 
    mutate(disease = str_replace(file_list2[i], pattern = ".assoc.stat.csv", replacement = ""))
  
}

sum.stat = map_dfr(stat_LIST, rbind) %>% 
  select("Normalised_Disease", "Assoc_Method", "Assoc_Alternative", 
"Assoc_Statistic", "Assoc_PValues", "proportion_cases_with_phers0", "disease") %>% 
  mutate(Assoc_P.FDR = p.adjust(Assoc_PValues, method = "BH")) %>% 
  left_join(sum.dis) %>% 
  mutate(Age_Classifier =ifelse(Median_age_of_onset==0, "At birth",
                         ifelse(Median_age_of_onset<18, "Before 18",
                         ifelse(Median_age_of_onset>60,"Over 60",
                         ifelse(Median_age_of_onset>40,"Over 40", "Between 18 and 40"))))) %>% 
    mutate(is_significant =ifelse(Assoc_P.FDR<0.05, "YES","NO")) 

```

## How many significant results?
```{r}
sum.stat %>% 
  ggplot(aes(is_significant,fill=is_significant)) +
  geom_bar(position = "dodge")
```

```{r}
sum.stat %>% 
  count(is_significant)
```


## Difference in age groups?
```{r}
sum.stat %>% 
  ggplot(aes(Age_Classifier,  fill=is_significant)) +
  geom_bar(position = "dodge")
```

# Calculate Statistical Power
```{r}
file_list = list.files("/home/smueller/health_records/smueller/phers/samples",
                        pattern="samples_ccratio_4.csv$")

effect_LIST = list()

for(i in 1:length(file_list)){
  
  dis = str_replace(file_list[i],pattern="_samples_ccratio_4.csv", replacement = "" )
  Normalised_Disease = str_replace_all(dis, pattern = "_", replacement =" ")

  samples = read.csv(paste0("~/health_records/smueller/phers/samples/",dis,"_samples_ccratio_4.csv")) %>% 
    select(participant_id, Pheno)
  
  phers= read.delim(paste0("~/health_records/smueller/phers/phers/",dis,".phers.txt")) %>% 
    select(1,2)
  
  samples_phers = samples %>%
    right_join(phers)
  
  effect_LIST[[i]] = wilcox_effsize(data=samples_phers,
                 formula= PRS~Pheno, 
                 alternative = "greater",
                 ci=T) %>% 
    mutate(Normalised_Disease=Normalised_Disease)
}

# combine
effect_size = map_dfr(effect_LIST,rbind)

# join with sum.stat and save
sum.stat2 = sum.stat %>% 
  left_join(effect_size %>%  select(Normalised_Disease, effsize, magnitude),
            by="Normalised_Disease" )

# save
write_csv(sum.stat2, 
          paste0(OUTPATH,"suppl_table.",SCRIPT,".summary_all_stats.",DATE,".csv"))

```

## Plot effect size against significane
```{r}
sum.stat2 %>% 
  ggplot(aes(-log10(Assoc_PValues), effsize, color= is_significant))+
  geom_point(size=2)
```

```{r}
sum.stat2 %>% 
  ggplot(aes(-log10(Assoc_PValues), magnitude))+
  geom_boxplot(fill=NA, outlier.colour = NA) +
  geom_point(aes(color=is_significant),size=2, position = position_jitter(height = 0.3))
```

