---
title: "Inspect_Cases_with_PRS_0"
author: "Stefanie Mueller"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
    lightbox: true
    gallery: true
    toc_depth: 3
    use_bookdown: true
    fig_caption: true
    code_folding: hide
editor_options: 
  chunk_output_type: console
---


```{r knitr_init, echo=FALSE, cache=FALSE}
# load libraries
library(knitr)
library(rmdformats)

## Global options
#options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=TRUE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE,
               fig.width = 10)
opts_knit$set(width=75)
```

# PREFACE
**PROJECT**:  Phenotype Risk Scores  
**OBJECTIVE**:  Inspect recruited participants with PheRS "0" for recruitment disease   
**DEPENDENCE**:  phers.R    
**REQUIREMENT**:  R 3.5.0   
<br>

## Load packages
```{r}
library(tidyverse)
library(purrr)
library(dominanceanalysis)
library(GGally)
library(ggpubr)
```

## Script Variables
```{r}
OUTPATH = "~/health_records/smueller/phers/tables/"
SCRIPT = "SC_Case0"
DATE = Sys.Date()
```

# Load data
```{r, eval=T}
#Phenotype Risk Scores per disease
phers_list = list.files("/home/smueller/health_records/smueller/phers/samples",
                        pattern="samples_ccratio_4.csv$")

phers_and_pheno_LIST = list()

#loop over diseases
for (i in 1:length(phers_list)){
  
  disease = str_replace(phers_list[i],pattern="_samples_ccratio_4.csv", replacement = "" ) 
  
  # read phers
  phers= read.delim(paste0("~/health_records/smueller/phers/phers/",disease,".phers.txt")) %>% 
    select(1,2)
  
  # read sample info
  phers_and_pheno_LIST[[i]] =   read.csv(paste0("/home/smueller/health_records/smueller/phers/samples/",(phers_list[i]))) %>% 
    mutate(disease = disease) %>% 
    left_join(phers) %>% 
    mutate(Type = ifelse(Pheno == "Control", "Control", ifelse(PRS==0, "Case0","Case")))
}

phers_and_pheno = map_dfr(phers_and_pheno_LIST, rbind)
```

# Make Summary Stat
## Per participants type in disease: ie. Case, Case0, Control
```{r, eval=F}
#summarise info per disease and particiapnt type (ie. control, case with PRS "0" or case)
sum.phers_and_pheno = phers_and_pheno %>% 
  group_by(disease, Type) %>% 
  summarise(Number =n(),
    Median_age = median(age),
    IQR_age = quantile(age,probs = 0.75)-quantile(age,probs = 0.25),
    Median_age_of_onset = median(age_of_onset, na.rm = T),
    IQR_age_of_onset = quantile(age_of_onset, probs =0.75,na.rm = T)- 
      quantile(age_of_onset, probs =0.75,na.rm = T),
    Median_encounters = median(Number_uniq_encoutner),
    IQR_encounters = quantile(Number_uniq_encoutner,probs = 0.75)-quantile(Number_uniq_encoutner,probs = 0.25),
    Median_Record_Length = median(Record_Length),
    IQR_Record_Length = quantile(Record_Length,probs = 0.75)-quantile(Record_Length,probs = 0.25),
    Median_Number_uniq_ICD = median(Number_uniq_ICD),
    IQR_Number_uniq_ICD = quantile(Number_uniq_ICD,probs = 0.75)-quantile(Number_uniq_ICD,probs = 0.25)) 


# save
write_csv(sum.phers_and_pheno, 
          paste0(OUTPATH,"suppl_table.",SCRIPT,".summary_stat.",DATE,".csv"))
```


```{r}
sum.phers_and_pheno= read.csv("~/health_records/smueller/phers/tables/suppl_table.SC_Case0.summary_stat.2020-10-13.csv")
```

## For cases in disease
```{r}
#summarise info per disease and particiapnt type (ie. control, case with PRS "0" or case)
sum.dis = phers_and_pheno %>% 
  filter(Type!="Control") %>% 
  group_by(disease) %>% 
  summarise(Number =n(),
    Median_age = median(age),
    IQR_age = quantile(age,probs = 0.75)-quantile(age,probs = 0.25),
    Median_age_of_onset = median(age_of_onset, na.rm = T),
    IQR_age_of_onset = quantile(age_of_onset, probs=0.75,na.rm = T) - 
      quantile(age_of_onset, probs=0.25,na.rm = T),
    Num_age_of_onset_missing = sum(is.na(age_of_onset)),
    Prop_age_of_onset_missing = Num_age_of_onset_missing/Number,
    Median_encounters = median(Number_uniq_encoutner),
    IQR_encounters = quantile(Number_uniq_encoutner,probs = 0.75)-quantile(Number_uniq_encoutner,probs = 0.25),
    Median_Record_Length = median(Record_Length),
    IQR_Record_Length = quantile(Record_Length,probs = 0.75)-quantile(Record_Length,probs = 0.25),
    Median_Number_uniq_ICD = median(Number_uniq_ICD),
    IQR_Number_uniq_ICD = quantile(Number_uniq_ICD,probs = 0.75)-quantile(Number_uniq_ICD,probs = 0.25)) 
```


# Logistic Model Case Type 
**Model**: Case0 ~ age + Number_uniq_ICD + Number_uniq_encoutner +Record_Length +age_of_onset  

```{r}
# define functions
LogReg_model = function(data){
  
  model = glm(data=data, family = "binomial",
      formula = CaseType ~ age + Number_uniq_ICD + Number_uniq_encoutner + Record_Length + age_of_onset)
  
  return(model)
}

summarise_LogReg_model = function(LogReg_model){
  
  out = data.frame(LogReg_Coef_Intercept = LogReg_model$coefficients[1],
                   LogReg_Coef_Age = LogReg_model$coefficients[2],
                   LogReg_Coef_Number_uniq_ICD = LogReg_model$coefficients[3],
                   LogReg_Coef_Number_uniq_encoutner = LogReg_model$coefficients[4],
                   LogReg_Coef_Record_Length = LogReg_model$coefficients[5],
                   LogReg_Coef_Age_of_onset = LogReg_model$coefficients[6],
                   LogReg_AIC = LogReg_model$aic,
                   LogReg_deviance_fit = LogReg_model$deviance,
                   LogReg_deviance_Null = LogReg_model$null.deviance)
  return(out)
}
```

```{r, eval=F}
# apply function per disease
glm_stats = phers_and_pheno %>% 
  # remove control samples
  filter(Type!="Control") %>% 
  # remap factors for Case0 and Case so that Case0 will be first value
  mutate(CaseType =factor(Type, levels = c("Case0","Case"))) %>% 
  group_by(disease) %>% 
  group_modify(~ summarise_LogReg_model(LogReg_model(.x)))

# save
write_csv(glm_stats, 
          paste0(OUTPATH,"suppl_table.",SCRIPT,".log_reg.",DATE,".csv"))
```

```{r}
glm_stats = read.csv("~/health_records/smueller/phers/tables/suppl_table.SC_Case0.log_reg.2020-10-14.csv")
```


# Varibale importance analysis
**Problem**: since variable sin logistice regression are not scaled feature importance can't be infered from regression coeffiecents   
**Method**: will use dominance analysis to estiamte feature importance of varables   
**Reference**: https://cran.r-project.org/web/packages/dominanceanalysis/vignettes/da-logistic-regression.html   

```{r}
# define function
summarise_variable_importance = function(LogReg_model){
  boot_dom_aglm = bootAverageDominanceAnalysis(LogReg_model,R=100)
  R2m = summary(boot_dom_aglm, fit.functions=c("r2.m"))
  
  out= data.frame(Boot100_General_Dominance_Age = R2m$r2.m$bs.E[1],
                  Boot100_General_Dominance_Number_uniq_ICD = R2m$r2.m$bs.E[2],
                  Boot100_General_Dominance_Number_uniq_encoutner = R2m$r2.m$bs.E[3],
                  Boot100_General_Dominance_Record_Length = R2m$r2.m$bs.E[4],
                  Boot100_General_Dominance_Age_of_onset = R2m$r2.m$bs.E[5])
  
  return(out)
}
```

```{r, eval=F}
# apply function per disease
glm_importance_stats = phers_and_pheno %>% 
  # remove control samples
  filter(Type!="Control") %>% 
  # remap factors for Case0 and Case so that Case0 will be first value
  mutate(CaseType =factor(Type, levels = c("Case0","Case"))) %>% 
  group_by(disease) %>% 
  group_modify(~ summarise_variable_importance(LogReg_model(.x)))

# save
write_csv(glm_importance_stats,
                    paste0(OUTPATH,"suppl_table.",SCRIPT,".feat_importance.",DATE,".csv"))
```

```{r}
glm_importance_stats = read.csv("~/health_records/smueller/phers/tables/suppl_table.SC_Case0.feat_importance.2020-10-14.csv")
```


# Plots
## Histogram of Proportion of Case0
Q: Can we see clustering in diseases or normal distribution?   
A: quite uniform distribution
```{r, eval=T}
options(bitmapType="cairo")
X11(type = 'cairo')

sum.phers_and_pheno %>% 
  select(disease, Type, Number) %>% 
  pivot_wider(id_cols = disease, names_from = Type, values_from = Number) %>% 
  mutate(Prop_Case0 = Case0/(Case+Case0)) %>% 
  ggplot(aes(Prop_Case0)) + 
  geom_histogram(binwidth = 0.01) +
  labs(title = "Histogram Proportion of Cases with PheRS == 0 per disease")
```

## Proportion of Case0 vs Median Age in disease
Q: Is there a correlation between median age in cases and proportion of cases with PheRS "0"?   
A: No
```{r}
options(bitmapType="cairo")
X11(type = 'cairo')

sum.phers_and_pheno %>% 
  select(disease, Type, Number, Median_age) %>% 
  pivot_wider(id_cols = disease, names_from = Type, values_from = Number) %>% 
  mutate(Prop_Case0 = Case0/(Case+Case0)) %>% 
  left_join(sum.dis) %>% 
  ggplot(aes(Prop_Case0, Median_age)) + 
  geom_point() +
  labs(title = "Proportion of Cases with PheRS == 0 vs Median age in Cases per disease") +
  geom_smooth(method="lm") +
  stat_cor(label.y = 65) +
  stat_regline_equation(label.y = 60)
```


