---
title: "missed_diagnoses"
author: "Stefanie Mueller"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
    lightbox: true
    gallery: true
    toc_depth: 3
    use_bookdown: true
    fig_caption: true
    code_folding: hide
editor_options: 
  chunk_output_type: console
---


```{r knitr_init, echo=FALSE, cache=FALSE}
# load libraries
library(knitr)
library(rmdformats)

# define functions
myDT = function(data){
  data %>% DT::datatable(rownames = FALSE,
    filter = 'bottom',
    extensions = 'Buttons',
      options = list(
        searching = TRUE,
        fixedColumns = TRUE,
        autoWidth = TRUE,
        dom = 'Bfrtip',
        buttons = c('copy', 'csv', 'excel'))
    )
}

## Global options
#options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=TRUE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE,
               fig.width = 10)
opts_knit$set(width=75)
```

# PREFACE
**PROJECT**:  Phenotype Risk Scores  
**OBJECTIVE**:  Identidy individuals with missed diagnoses   
**DEPENDENCE**:  phers.R    
**REQUIREMENT**:  R 3.5.0   
<br>

## Load packages
```{r}
library(tidyverse)
library(purrr)
```

## Script Variables
```{r}
OUTPATH = "~/health_records/smueller/phers/tables/"
SCRIPT = "MissDiag"
DATE = Sys.Date()
```

# Load PheRS for and case/control status
```{r, eval=T}
#Phenotype Risk Scores per disease
phers_list = list.files("/home/smueller/health_records/smueller/phers/samples",
                        pattern="samples_ccratio_4.csv$")

phers_and_pheno_LIST = list()

#loop over diseases
for (i in 1:length(phers_list)){
  
  disease = str_replace(phers_list[i],pattern="_samples_ccratio_4.csv", replacement = "" ) 
  
  # read phers
  phers= read.delim(paste0("~/health_records/smueller/phers/phers/",disease,".phers.txt")) %>% 
    select(1,2)
  
  # read sample info
  phers_and_pheno_LIST[[i]] =   read.csv(paste0("/home/smueller/health_records/smueller/phers/samples/",(phers_list[i]))) %>% 
    mutate(disease = disease) %>% 
    right_join(phers) 
}

```

# Find threshold of 75% percentile in cases
```{r}
# function to get 75% percentile
get_threshold_case_perc75 = function(data){
  
  disease = data$disease[1]
  
  thresh = data %>% 
    filter(isCase) %>% 
    select(PRS) %>% 
    unlist() %>% 
    quantile(0.75)
  
  out =data.frame(disease = disease,
                  threshold = thresh)
  return(out)
  
}

case_perc75 = function(data){
  
  thresh = data %>% 
    filter(isCase) %>% 
    select(PRS) %>% 
    unlist() %>% 
    quantile(0.75)
  
  exceeding = data %>%
    filter(!isCase) %>% 
    filter(PRS>= thresh) 
  
  return(exceeding)
  
}

threshold_case_perc75_LIST = map(phers_and_pheno_LIST, ~ get_threshold_case_perc75(.x))

controls_exceeding_c75_LIST = map(phers_and_pheno_LIST, ~ case_perc75(.x))

threshold_case_perc75= map_dfr(threshold_case_perc75_LIST, rbind) %>% 
    mutate(NonCaseExceeding = lapply(controls_exceeding_c75_LIST,nrow) %>% unlist()) 

controls_exceeding_c75 = map_dfr(controls_exceeding_c75_LIST, rbind)

```

## Save
```{r}
write_csv(threshold_case_perc75, 
          paste0(OUTPATH,"suppl_table.",SCRIPT,".case_q75_threshold.",DATE,".csv"))
write_csv(controls_exceeding_c75, 
          paste0(OUTPATH,"suppl_table.",SCRIPT,".case_q75_threshold_exceeding_samples.",DATE,".csv"))
```

## what is the thrshold and how many controls have PheRS above it
```{r}
threshold_case_perc75 %>% 
  myDT()
```

## plot some info
```{r}
threshold_case_perc75= read.csv("~/health_records/smueller/phers/tables/suppl_table.MissDiag.case_q75_threshold.2020-10-22.csv")
controls_exceeding_c75 = read.csv("~/health_records/smueller/phers/tables/suppl_table.MissDiag.case_q75_threshold_exceeding_samples.2020-10-22.csv")
sum.stat = read.csv("~/health_records/smueller/phers/tables/suppl_table.DisOver.summary_all_stats.2020-10-26.csv")

```

### Histogram of threshold for identification of closer inspetion of Non-cases
 
removed disease with non significant result
```{r}
options(bitmapType="cairo")
X11(type = 'cairo')

controls_exceeding_c75 %>% 
  group_by(disease) %>% 
  summarise(Number=n()) %>% 
  left_join(threshold_case_perc75) %>% 
  left_join(sum.stat, by="disease") %>% 
  filter(is_significant=="YES") %>% 
  ggplot(aes(threshold)) +
  geom_histogram()+
  labs(x="Threshold \n75% Percentile Cases")
```


### Number of Non-Cases above threshold vs threshold straified my effect size magnitude

removed disease with non significant result
```{r}
controls_exceeding_c75 %>% 
  group_by(disease) %>% 
  summarise(Number=n()) %>% 
  left_join(threshold_case_perc75) %>% 
  left_join(sum.stat, by="disease") %>% 
  filter(is_significant=="YES") %>% 
  ggplot(aes(threshold,Number.x)) +
  geom_point(aes(color=magnitude), size=2)+
  scale_y_log10(labels=c(1,3,10,30,100,300,1000,3000),
                breaks=c(1,3,10,30,100,300,1000,3000)) +
  labs(x="Threshold \n75% Percentile Cases", y="Number Non-Cases with\n PheRS above threshold\n (log Scale)")
```

